Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States.

### 2025-05-02 products/manuscript/Manuscript.qmd
Changes for final product

---
title: "Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States."
author: "Shaun van den Hurk"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
#install.packages("magick")
library(here)
library(knitr)
library(magick)
```

@@

-   First Author: Shaun van den Hurk$^{1,2,*}$

@@

$*$ These authors contributed equally to this work.

@@

Cardiovascular disease (CVD) remains a leading cause of mortality in the United States, with hypertension recognized as a major modifiable contributor. This study investigates patterns and predictors of hypertension-related CVD mortality among U.S. adults aged 35 and older using county-level data from 2000 to 2019. We explored temporal, demographic, and geographic disparities and evaluated the predictive performance of demographic and temporal factors using machine learning models.
Descriptive analyses revealed persistent and pronounced disparities in mortality across race, sex, and age groups. Black and American Indian/Alaska Native populations experienced the highest mortality rates, particularly among adults aged 65 and older. Temporal modeling showed diverging trends across subgroups, with Black adults seeing the steepest declines, while mortality increased among males and American Indian/Alaska Native populations. Spatial clustering analyses identified modest but statistically significant autocorrelation, with localized high-mortality clusters in the Southeastern U.S.
Finally, predictive modeling using Random Forest and LASSO regression demonstrated that demographic and temporal variables could explain a substantial proportion of county-level variation in mortality. The Random Forest model achieved an R² of 0.748 and identified race/ethnicity and year as the most influential predictors. These findings highlight the enduring influence of demographic and geographic factors on cardiovascular health outcomes and support the use of predictive models to inform future public health strategies.

@@

Lifestyle diseases such as cardiovascular disease (CVD) represent some
of the most significant causes of death globally, both in low-income
regions and high-income countries such as the United States of America
(USA). Cardiovascular diseases are a broad umbrella term including
different clinical manifestations such as hypertension, stroke, and
ischemic heart disease. Cardiovascular diseases were cited as the
leading cause of death globally in 2022 (..1..), in the USA they were
responsible for one in every five deaths nationally (...2.).
Hypertension is commonly identified as a significant risk factor for
multiple cardiovascular conditions, and one that is modifiable, and thus
represents a good metric for potential public health interventions
(...3.). However, not all demographics appear to be equally affected by
hypertension and CVD, with differences in rates of CVD and hypertension
in people from different races, sexes, age groups, geographic location,
income levels, and other groups in previous studies (..4..). A deeper
understanding of these factors helps us to get a deeper understanding of
the multifactorial nature of CVD and also helps guide effective public
health interventions. Additionally, it is important to understand if
there have been any differences in these rates of CVD and hypertension
over time, and if there are particular geographic hotspots. This is
essential as it allows public health officials to measure the impact of
previous intervention strategies and to ensure that resources
strategically target areas of concern.

@@

The dataset was collected from the CDC website and the information was
sourced from the National Vital Statistics System (..5...). The dataset
contains rates and trends in the hypertension-related cardiovascular
disease (CVD) death rates. The study period is from 2000 to 2019, with
two main intervals displayed in the data 2000-2009, and 2010-2019. The
data is further distributed by county, age group (35-64 years and 65
years and older), race/ethnicity, and sex.

(1= )

(2= Virani SS, et al. Heart Disease and Stroke Statistics—2020 Update: A
Report From the American Heart Association. *Circulation*.
2020;141:e139–e596.)

(3= Whelton PK, et al. 2017 ACC/AHA Guideline for the Prevention,
Detection, Evaluation, and Management of High Blood Pressure in Adults.
*Hypertension*. 2018;71:e13–e115)

(Mahmood SS, et al. The Framingham Heart Study and the epidemiology of
cardiovascular disease: a historical perspective. *Lancet*.
2014;383(9921):999–1008.)

(4= Yusuf S, et al. Modifiable risk factors, cardiovascular disease, and
mortality in 155,722 individuals from 21 high-income, middle-income, and
low-income countries (PURE): a prospective cohort study. *Lancet*.
2020;395(10226):795–808.)

(undefined

(5= at the following url:
https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
)

@@

\- Temporal trends differ significantly across race and sex, with
widening inequities over time.

\- Predictive models can identify key demographic and spatial factors
associated with higher mortality trends.

@@

We obtained the dataset from the CDC’s online portal, sourced from the National Vital Statistics System [5]. The raw CSV file was approximately 300MB in size and contained 1,103,872 rows and 24 columns. Initial preprocessing was conducted in R to remove redundant variables, reduce file size, and prepare the dataset for analysis.There were some initial processing steps performed to evaluate and delete some of the unnecessary columns and to save this data to an RDS
document to reduce the size. This was all outlined in the "processingcode" document (and README file), which will not run
automatically because the Excel (csv) file will not be available due to
size concerns with Github uploads (available on request). This process is fully documented in the processingcode.R and processingfile-v1.qmd files within the project repository.

@@

All rows with missing values in any of the key variables (mortality rate, year, age group, race/ethnicity, sex, and geographic coordinates) were removed. The resulting cleaned dataset contained 572,572 rows and 14 variables. Summary statistics and exploratory inspection (e.g., using the skimr package) confirmed the data's readiness for analysis. The dataset included both numeric variables (e.g., mortality rates, year, coordinates) and categorical variables (e.g., sex, race/ethnicity, age group, county).

@@

All cleaning and preprocessing steps are reproducible and are logged in the repository.

@@

For spatial analysis (Aim 2), we examined county-level clustering of mortality rates using spatial statistics. County geometries were created by converting longitude and latitude coordinates into spatial features using the sf package. Spatial weights were computed using 5-nearest-neighbor relationships. Global spatial autocorrelation was assessed using Moran’s I statistic. Local clustering patterns were evaluated using Local Indicators of Spatial Association (LISA) via the spdep package. Local spatial relationships were visualized using standardized mortality z-scores and classified into high–high, low–low, or spatial outlier clusters based on the combination of mortality values and neighboring spatial lags.

To evaluate demographic disparities in temporal trends (Aim 3), we extended the GLMs used in Aim 1 by including three-way interactions between year, race/ethnicity, and sex, and stratified analyses by age group. Estimated marginal trends were obtained separately for adults aged 35–64 and those aged 65 and older. Annual slope estimates and their confidence intervals were extracted and compared across race–sex combinations. Pairwise comparisons of slopes were again conducted using emmeans, with the goal of quantifying and visualizing disparities in mortality trends across key demographic intersections.

In the final aim (Aim 4), we developed predictive models to estimate county-level mortality rates using demographic and temporal features. Two modeling approaches were implemented: a random forest model using the ranger engine and a LASSO regression model using the glmnet engine. The modeling dataset included year, age group, race/ethnicity, and sex as predictors. Categorical variables were dummy-encoded using the recipes package, and numeric predictors were normalized; zero-variance predictors were removed prior to modeling. Data were split into training (80%) and test (20%) sets, stratified by mortality outcome. Within the training set, 10-fold cross-validation was used to tune model hyperparameters. For random forests, tuning was performed over a regular grid of mtry and min_n values. LASSO regression was tuned using a log-scaled penalty grid to identify optimal shrinkage parameters. Final model performance was evaluated on the test set using RMSE (root mean squared error), MAE (mean absolute error), and R² (coefficient of determination). Confidence intervals for these metrics were not computed due to computational constraints. Model interpretability was further assessed using variable importance plots for the random forest model and non-zero coefficient selection in LASSO.

@@

## Descriptive and Exploratory Analysis
We began by examining overall patterns in hypertension-related cardiovascular disease (CVD) mortality among U.S. adults aged 35 and older from 2000 to 2019 (Aim 1; Hypothesis 1). Summary statistics revealed large disparities in mortality rates across racial/ethnic and age groups. For adults aged 35–64, Black individuals had the highest mean county-level mortality rate (92.2 per 100,000), followed by American Indian/Alaska Native (46.2), and White (32.9). In the 65+ age group, Black adults again had the highest mean mortality (662.6), with American Indian/Alaska Native (451.9) and White (398.3) also experiencing elevated rates.

Boxplots of mortality by race/ethnicity and sex (Figures below) confirmed these disparities, showing right-skewed distributions with particularly high upper ranges for Black and American Indian/Alaska Native adults. Males had consistently higher mortality distributions than females across all groups.

Temporal visualizations indicated divergent patterns between subgroups. For example, while some populations experienced declines in mortality rates, others showed upward or stable trends. These descriptive findings supported formal modeling to assess temporal and demographic patterns.

```{r, echo=FALSE, message=FALSE}
mortality_race_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_race.png"))
print(mortality_race_boxplot)

mortality_sex_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_sex.png"))
print(mortality_sex_boxplot)
```

@@

### Temporal Trends in Mortality 
At the national level, a linear regression of overall mortality against year showed a small but statistically significant increase in mortality from 2000 to 2019 (slope = 1.40, p < 0.001). However, stratified models revealed considerable variation between groups.

A generalized linear model (GLM) incorporating race/ethnicity and sex predictors (Aim 1.2; Hypotheses 1 and 3) confirmed that Black individuals had the highest predicted mortality rates, followed by White and American Indian/Alaska Native populations. Males had significantly higher predicted rates than females across all racial/ethnic groups.

An interaction model between year and race (Aim 1.3) revealed statistically significant divergence in trends. Mortality among Black individuals declined sharply over time (slope = –7.06, p < 0.001), while American Indian/Alaska Native (slope = –3.41) and Hispanic populations showed shallower or increasing trends.

Interaction between year and age group (Aim 1.4) demonstrated that adults aged 65+ experienced a significantly steeper increase in mortality over time compared to those aged 35–64 (interaction slope = +0.22, p < 0.001). Estimated marginal trends (Aim 1.5) further confirmed this: the slope among older adults was 1.26 per year (95% CI: 1.18–1.34), versus 1.04 (95% CI: 0.96–1.12) in the younger group (p = 0.0002).

```{r, echo=FALSE, message=FALSE}
#Print the glm plot
glm_interaction_plot <- image_read(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"))
print(glm_interaction_plot)
```

### Geographic Distribution and Clustering 
Spatial visualizations demonstrated considerable geographic variation in mortality by race/ethnicity (Aim 2.1–2.2). Elevated mortality was most prominent among Black individuals, particularly in counties across the Southeast and Midwest. Asian and Pacific Islander populations displayed lower and more geographically diffuse mortality rates.

```{r, echo=FALSE, message=FALSE}
combined_spatial_plot <- image_read(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"))
print(combined_spatial_plot)
```

Global spatial clustering was modest but statistically significant for the national dataset (Moran’s I = 0.0435, p < 0.001), with similar low levels of spatial autocorrelation in state-level analyses: Georgia (I = 0.0297), Mississippi (0.0289), Texas (0.0168), and Kentucky (0.0227), all p < 0.001 (Aim 2.3).

Local spatial clustering (Aim 2.4) was visualized using Local Indicators of Spatial Association (LISA). Statistically significant high–high clusters were identified in counties across parts of the Southeast, but the majority of counties showed no local spatial clustering. This suggests that while some mortality hotspots exist, broad demographic and structural drivers likely contribute more to the observed geographic distribution than tightly bounded spatial clusters.

### Demographic Disparities in Temporal Trends
We assessed demographic disparities in mortality trends across race/ethnicity and sex within age groups (Aim 3.1–3.3; Hypothesis 3). Among adults aged 35–64, Black individuals experienced the steepest annual decline in mortality (–3.68), while American Indian/Alaska Native individuals showed the largest increases (+3.38). Overall, males experienced steeper increases than females.

Pairwise comparisons showed the greatest differences between Overall Male and Black Overall groups (+7.00 per year), followed by contrasts involving American Indian/Alaska Native, White, and Asian/Pacific Islander groups. These results indicate persistent and widening disparities in mortality trends, particularly among marginalized populations.

```{r, echo=FALSE, message=FALSE}
readRDS(here::here("results", "tables", "slope_table_agegroup_fmt.rds"))
readRDS(here::here("results", "tables", "slope_table_race_sex_fmt.rds"))
```

Additional visualizations from Aim 3.4 reinforce these findings:

```{r, echo=FALSE, message=FALSE}
slope_plot_35_64 <- image_read(here::here("results", "figures", "slope_plot_35_64.png"))
print(slope_plot_35_64)

slope_plot_65_plus <- image_read(here::here("results", "figures", "slope_plot_65_plus.png"))
print(slope_plot_65_plus)

@@

### Predictive Modeling of Mortality Rates

To evaluate the ability of demographic and temporal factors to predict hypertension-related CVD mortality at the county level, we developed and compared two supervised machine learning models: a Random Forest and a LASSO regression model (Aim 4; Hypothesis 4). Both models were trained on 80% of the dataset and evaluated on a held-out 20% test set, with 10-fold cross-validation used during training to optimize hyperparameters and reduce overfitting.

The Random Forest model demonstrated strong predictive accuracy. The best-performing model used all four available predictors (year, age group, race/ethnicity, and sex) with a minimum node size of 15. On the test set, this model achieved a root mean squared error (RMSE) of 115.3, a mean absolute error (MAE) of 66.9, and an R² of 0.748, indicating that approximately 75% of the variance in county-level mortality was explained by the model. Variable importance scores revealed that race/ethnicity and year were the most influential predictors, followed by age group and sex. The predicted versus observed plot showed generally good alignment, though some dispersion at the extremes suggested modest underfitting in counties with very high or low mortality.

```{r, echo=FALSE, message=FALSE}
model_performance_summary_table <- readRDS(here::here("results", "tables", "model_performance_summary.rds"))
model_performance_summary_table

@@

```{r, echo=FALSE, message=FALSE}
rf_pred_plot <- image_read(here::here("results", "figures", "rf_pred_vs_obs.png"))
print(rf_pred_plot)

rf_vip_plot <- image_read(here::here("results", "figures", "rf_variable_importance.png"))
print(rf_vip_plot)

@@

The LASSO regression model provided a more interpretable, though slightly less accurate, alternative. On the same test set, the LASSO model yielded an RMSE of 119.5, an MAE of 70.7, and an R² of 0.726. This model performed coefficient shrinkage and variable selection, shrinking several dummy-coded predictors toward zero. While this feature enhances interpretability and highlights influential variables, it also limits the model’s flexibility to capture complex non-linear patterns. The predicted versus observed plot for the LASSO model revealed horizontal banding patterns. This reflects the impact of dummy-encoded categorical variables — many counties shared identical predictor combinations, resulting in repeated predictions and horizontal clustering in plots. These effects were more pronounced in LASSO than in the Random Forest, which is better able to handle categorical interactions.

```{r, echo=FALSE, message=FALSE}
lasso_pred_plot <- image_read(here::here("results", "figures", "lasso_pred_vs_obs.png"))
print(lasso_pred_plot)

@@

Despite these differences, both models demonstrated solid performance in capturing patterns in the data. The Random Forest provided better predictive accuracy, while the LASSO model offered greater clarity in identifying relevant features. Taken together, these results indicate that demographic and temporal factors are strong predictors of mortality variation at the county level, supporting Hypothesis 4 and offering tools for future public health risk modeling.

>>>>

---
title: "Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States."
author: "Shaun van den Hurk"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/references/CVD-project-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load a few R packages
#install.packages("magick")
library(here)
library(knitr)
library(magick)
```

@@

-   First Author: Shaun van den Hurk$^{1*}$

@@

@@

Cardiovascular disease (CVD) remains a leading cause of mortality in the United States, with hypertension recognized as a major modifiable contributor. This study investigates patterns and predictors of hypertension-related CVD mortality among U.S. adults aged 35 and older using county-level data from 2000 to 2019. We examined temporal, demographic, and geographic disparities to assess both descriptive trends and predictive capacity.

Exploratory analyses revealed persistent disparities in mortality across race, sex, and age groups. Black and American Indian/Alaska Native adults consistently exhibited the highest mortality rates, especially among those aged 65 and older. Temporal models showed diverging trends across subgroups, with steep declines in Black adults and rising mortality among males and American Indian/Alaska Native populations. Spatial clustering was modest but statistically significant, with localized high-mortality areas concentrated in the Southeast.

Predictive modeling using Random Forest and LASSO regression demonstrated that demographic and temporal variables explained a substantial proportion of county-level mortality variation. The Random Forest model achieved an R² of 0.748, with age group emerging as the most influential predictor, followed by race/ethnicity. These findings underscore the enduring impact of demographic and geographic factors on cardiovascular health outcomes and support the use of predictive models to inform targeted public health interventions.

@@

Lifestyle diseases such as cardiovascular disease (CVD) represent some
of the most significant causes of death globally, both in low-income
regions and high-income countries such as the United States of America
(USA). Cardiovascular diseases are a broad umbrella term including
different clinical manifestations such as hypertension, stroke, and
ischemic heart disease. Cardiovascular diseases were cited as the
leading cause of death globally in 2022, in the USA they were
responsible for one in every five deaths nationally [@mills_global_2020; @virani_heart_2020; @vaughan_countylevel_2022].
Hypertension is commonly identified as a significant risk factor for
multiple cardiovascular conditions, and one that is modifiable, making it an ideal target for public health interventions [@whelton_2017_2018; @mahmood_framingham_2014]. However, not all demographics appear to be equally affected by
hypertension and CVD, with differences in rates of CVD and hypertension
in people from different races, sexes, age groups, geographic location,
income levels, and other groups in previous studies [@mahmood_framingham_2014; @yusuf_modifiable_2020].

A deeper understanding of these factors helps us to address the multifactorial nature of CVD and also helps guide effective public health interventions. 
Additionally, it is important to understand if
there have been any differences in these rates of CVD and hypertension
over time, and if there are particular geographic hotspots. This is
essential as it allows public health officials to measure the impact of
previous intervention strategies and to ensure that resources
strategically target areas of concern.

@@

The dataset was collected from the CDC website and the information was
sourced from the National Vital Statistics System [@noauthor_rates_nodate]. The dataset
contains rates and trends in the hypertension-related cardiovascular
disease (CVD) death rates. The study period is from 2000 to 2019, with
two main intervals displayed in the data 2000-2009, and 2010-2019. The
data is further distributed by county, age group (35-64 years and 65
years and older), race/ethnicity, and sex.

@@

\- Temporal trends differ across race and sex, with widening inequities over time.

\- Predictive models can identify key demographic and spatial features associated with elevated mortality risk.

@@

We obtained the dataset from the CDC’s online portal, sourced from the National Vital Statistics System [@noauthor_rates_nodate]. The raw CSV file was approximately 300MB in size and contained 1,103,872 rows and 24 columns. Initial preprocessing was conducted in R to remove redundant variables, reduce file size, and prepare the dataset for analysis. This was all outlined in the "processingcode" document (and README file), which will not run
automatically because the Excel (csv) file will not be available due to
size concerns with uploading documents (available on request). Full preprocessing steps are documented in the repository’s processingfile-v1.qmd and processingcode.R scripts.

@@

All rows with missing values in any of the key variables (mortality rate, year, age group, race/ethnicity, sex, and geographic coordinates) were removed. The resulting cleaned dataset contained 572,572 rows and 14 variables. Summary statistics and exploratory inspection (e.g. using the skimr package) confirmed the data's readiness for analysis. The dataset included both numeric variables (e.g. mortality rates, year, coordinates) and categorical variables (e.g. sex, race/ethnicity, age group, county).

@@

@@

For spatial analysis (Aim 2), we examined county-level clustering of mortality rates using spatial statistics. County geometries were created by converting longitude and latitude coordinates into spatial features using the sf package. Spatial weights were computed using 5-nearest-neighbor relationships. Global spatial autocorrelation was assessed using Moran’s I statistic. Local clustering patterns were evaluated using Local Indicators of Spatial Association (LISA) via the spdep package. We visualized local spatial relationships using standardized mortality z-scores, classifying counties as high–high, low–low, or spatial outliers depending on both their mortality values and those of neighboring counties.

To evaluate demographic disparities in temporal trends (Aim 3), we extended the GLMs used in Aim 1 by including three-way interaction terms between year, race/ethnicity, and sex, and stratified analyses by age group. Estimated marginal trends were obtained separately for adults aged 35–64 and those aged 65 and older. Annual slope estimates and their confidence intervals were extracted and compared across race–sex combinations. Pairwise slope comparisons were again conducted using emmeans, allowing us to quantify and visualize disparities across key demographic intersections.

In the final aim (Aim 4), we developed predictive models to estimate county-level mortality rates using demographic and temporal features. Two modeling approaches were implemented: a random forest model using the ranger engine and a LASSO regression model using the glmnet engine. The modeling dataset included year, age group, race/ethnicity, and sex as predictors. Categorical variables were dummy-encoded using the recipes package, and numeric predictors were normalized; zero-variance predictors were removed. Data were split into training (80%) and test (20%) sets, stratified by mortality outcome. Within the training set, 10-fold cross-validation was used to tune model hyperparameters. For random forests, tuning was performed over a regular grid of mtry and min_n values. LASSO regression was tuned using a log-scaled penalty grid to identify optimal shrinkage parameters. Final model performance was evaluated on the test set using RMSE (root mean squared error), MAE (mean absolute error), and R² (coefficient of determination). Confidence intervals were computed for cross-validation metrics but not for the final test set performance due to computational constraints. Model interpretability was further assessed using variable importance plots for the random forest model and non-zero coefficient selection in LASSO.

@@

## Descriptive and Exploratory Analysis
We began by examining overall patterns in hypertension-related cardiovascular disease (CVD) mortality among U.S. adults aged 35 and older from 2000 to 2019. Summary statistics revealed large disparities in mortality rates across racial/ethnic and age groups. For adults aged 35–64, Black individuals had the highest mean county-level mortality rate (92.2 per 100,000), followed by American Indian/Alaska Native (46.2), and White (32.9). In the 65+ age group, Black adults again had the highest mean mortality (662.6), with American Indian/Alaska Native (451.9) and White (398.3) also experiencing elevated rates. 
To further characterize these disparities, we summarized the mean and median mortality rates by race/ethnicity across all counties and years. Black adults (all ages) had the highest mean mortality (305 per 100,000), while White adults (all ages) had the highest median mortality (185 per 100,000), reflecting differing distributions and the influence of extreme values (see Supplementary Table S1). These differences are visualized in boxplots by race/ethnicity and sex (@fig-mortality-rate-by-race and @fig-mortality-rate-by-sex).

Boxplots confirmed these disparities, showing right-skewed distributions with particularly high upper ranges for Black and American Indian/Alaska Native adults. Males consistently had higher mortality distributions than females across all groups. While some groups showed declines in mortality over time, others experienced stable or rising trends, supporting the need for formal modeling to assess temporal and demographic patterns

Temporal visualizations indicated divergent patterns between subgroups in @fig-glm-interaction. For example, while some populations experienced declines in mortality rates, others showed upward or stable trends. These descriptive findings supported formal modeling to assess temporal and demographic patterns.

```{r, echo=FALSE, message=FALSE}
#| label: fig-mortality-rate-by-race
#| fig-cap: "County-Level Mortality Rates by Race/Ethnicity (2000–2019). Boxplot shows distribution across counties and years."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "mortality_by_race.png"))

```{r, echo=FALSE, message=FALSE}
#| label: fig-mortality-rate-by-sex
#| fig-cap: "County-Level Mortality Rates by Sex (2000–2019). Boxplot shows distribution across counties and years."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "mortality_by_sex.png"))
```

@@

### Temporal Trends in Mortality 
At the national level, a linear regression of overall mortality against year showed a small but statistically significant increase in mortality from 2000 to 2019 (slope = 1.40, p < 0.001). However, stratified models revealed considerable variation between groups which may otherwise be masked by this overall increase.

To better capture these differences, we fit a generalized linear model (GLM) including an interaction between year and race/ethnicity. The model revealed substantial divergence in mortality trends across groups. Black adults had the highest predicted mortality in 2000 but experienced a steep decline over time (slope = –7.06, p < 0.001), while mortality rates increased among American Indian/Alaska Native adults. White adults showed a modest rise in mortality over the twenty year period. These patterns and contrasts were not evident in a simpler main-effects-only model, which is presented in Supplementary Figure S1 for comparison.

@fig-glm-interaction shows observed county-level mean mortality rates overlaid with GLM-predicted trends by group.

```{r, echo=FALSE, message=FALSE}
#| label: fig-glm-interaction
#| fig-cap: "Observed and model-predicted mortality rates by race/ethnicity (2000–2019), based on a generalized linear model with interaction between year and race/ethnicity. Dots represent annual observed means; lines show GLM predictions adjusting for sex."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "glm_interaction_plot.png"))
```

Building on the race/ethnicity model, we next assessed differences in mortality trends by age group using an interaction between year and age. This analysis revealed that adults aged 65 and older experienced a significantly steeper increase in mortality over time than those aged 35–64 (interaction slope = +0.22, p < 0.001). Estimated marginal trends from the model further supported this finding: the slope for older adults was 1.26 deaths per 100,000 per year (95% CI: 1.18–1.34), compared to 1.04 (95% CI: 0.96–1.12) among younger adults, a difference that was statistically significant (p = 0.0002). These results are visualized in Figure @fig-slope-agegroup, which presents the estimated annual mortality increase for each age group with 95% confidence intervals.

```{r, echo=FALSE, message=FALSE}
#| label: fig-slope-agegroup
#| fig-cap: "Estimated annual increase in hypertension-related mortality by age group. Slope values represent the yearly increase in predicted mortality rates per 100,000, based on a GLM with interaction between year and age group. Error bars reflect 95% confidence intervals."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "slope_plot_age_group.png"))
```

A corresponding figure from the GLM without interaction terms is available in the Supplement (Supplementary Figure S1). It assumes parallel trends across groups and does not capture the observed divergence over time.

### Geographic Distribution and Clustering 
@fig-spatial-all, @fig-spatial-black, and @fig-spatial-asian show spatial variation in hypertension-related mortality across all U.S. counties, both overall and for selected racial/ethnic groups. Elevated mortality was most prominent among Black individuals, particularly in counties across the Southeast and Midwest. Asian and Pacific Islander populations displayed lower and more geographically diffuse mortality patterns. Additional subgroup maps are included in the Supplement (Figures S2 - S6), and the combined spatial overview plot (Supplementary Figure S3) provides a side-by-side comparison of geographic patterns by group.

```{r, echo=FALSE, message=FALSE}
#| label: fig-spatial-all
#| fig-cap: "Spatial distribution of hypertension-related mortality – All races."
#| fig-width: 6.5
#| fig-height: 4.5
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "spatial_plot_all.png"))
```

```{r, echo=FALSE, message=FALSE}
#| label: fig-spatial-black
#| fig-cap: "Spatial distribution of hypertension-related mortality – Black population."
#| fig-width: 6.5
#| fig-height: 4.5
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "black_spatial_plot.png"))
```

```{r, echo=FALSE, message=FALSE}
#| label: fig-spatial-asian
#| fig-cap: "Spatial distribution of hypertension-related mortality – Asian and Pacific Islander population."
#| fig-width: 6.5
#| fig-height: 4.5
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "asian_spatial_plot.png"))
```

Global spatial clustering of hypertension-related CVD mortality was modest but statistically significant at the national level, with Moran’s I = 0.0435 (p < 0.001). Similar low levels of spatial autocorrelation were observed in state-level analyses: Georgia (I = 0.0297), Mississippi (0.0289), Texas (0.0168), and Kentucky (0.0227), all p < 0.001. To further explore local patterns, we conducted Local Indicators of Spatial Association (LISA) analyses. Supplementary Figure S7 displays the LISA cluster map for all U.S. counties, revealing several high–high clusters concentrated in the Southeastern United States, particularly in regions with high Black mortality. Supplementary Figures S8 and S9 provide state-level LISA maps for Georgia and Kentucky, which further illustrate the localized nature of spatial clustering. Although some statistically significant clusters were observed, the majority of counties did not exhibit strong local spatial dependence. These findings suggest that while certain mortality hotspots exist, broader structural and demographic factors likely account for the majority of geographic variation, rather than tightly bounded spatial processes.

### Demographic Disparities in Temporal Trends
We assessed demographic disparities in mortality trends by analyzing race/ethnicity and sex within each age group using GLMs with Year × Race × Sex interaction terms. Among adults aged 35–64, Black individuals experienced the steepest annual decline in mortality (–3.68 per 100,000), while American Indian/Alaska Native individuals showed the greatest increases (+3.38). Similar patterns were observed in the 65+ group, though the absolute changes were larger. Overall, males experienced steeper annual increases in mortality compared to females. Pairwise slope comparisons revealed the greatest disparity between Overall Male and Black Overall groups (+7.00 per year), followed by contrasts involving American Indian/Alaska Native, White, and Asian/Pacific Islander populations. These trends are summarized in @tbl-slope-race-sex and visualized in @fig-slope-35-64 and @fig-slope-65-plus. These results indicate persistent and widening disparities in hypertension-related mortality trends, particularly among historically marginalized populations, underscoring the need for targeted public health strategies.

```{r}
#| label: tbl-slope-agegroup
#| tbl-cap: "Estimated annual increase in hypertension-related mortality by age group. Values are based on a generalized linear model with an interaction between year and age group. Slopes represent predicted yearly change in mortality rates per 100,000."
#| echo: FALSE
knitr::include_graphics(here::here("results", "tables", "slope_table_agegroup.png"))

@@

```{r}
#| label: tbl-slope-race-sex
#| tbl-cap: "Estimated annual mortality trends by race/ethnicity and sex. Slopes were derived from a generalized linear model with Year × Race × Sex interaction terms and represent predicted yearly change in mortality per 100,000. Confidence intervals are shown in the final column."
#| echo: FALSE
knitr::include_graphics(here::here("results", "tables", "slope_table_race_sex.png"))

@@

Additional visualizations from Aim 3.4 reinforce these findings (@fig-slope-35-64 and @fig-slope-65-plus):

```{r, echo=FALSE, message=FALSE}
#| label: fig-slope-35-64
#| fig-cap: "Estimated annual change in hypertension-related mortality for adults aged 35–64, by race and sex. Values represent GLM-estimated slopes with 95% confidence intervals. 'Overall Male' and 'Overall Female' refer to aggregate trends across all racial/ethnic groups within the respective sex category."
#| echo: false
knitr::include_graphics(here::here("results", "figures", "slope_plot_35_64.png"))
```

```{r}
#| label: fig-slope-65-plus
#| fig-cap: "Estimated annual change in hypertension-related mortality for adults aged 65 and older, by race and sex. Values represent GLM-estimated slopes with 95% confidence intervals. 'Overall Male' and 'Overall Female' refer to aggregate trends across all racial/ethnic groups within the respective sex category."
#| echo: false
knitr::include_graphics(here::here("results", "figures", "slope_plot_65_plus.png"))
```

### Predictive Modeling of Mortality Rates 
To evaluate the predictive value of demographic and temporal factors for hypertension-related CVD mortality at the county level, we developed two supervised machine learning models: a Random Forest and a LASSO regression model. Each model was trained on 80% of the data and evaluated on a 20% held-out test set, with 10-fold cross-validation used to optimize hyperparameters and minimize overfitting.

The Random Forest model achieved the highest predictive accuracy. The optimal model used all four predictors—year, age group, race/ethnicity, and sex—with a minimum node size of 15 and mtry = 4. On the test set, it produced an RMSE of 115.3, MAE of 66.9, and R² of 0.748, explaining approximately 75% of the variance in county-level mortality (see @tbl-model-performance). Variable importance scores, calculated using permutation-based methods, highlighted age group (particularly 65+) as the most influential predictor, followed by race/ethnicity. In contrast, sex and year had lower relative importance (@fig-rf-vip). The predicted versus observed plot (@fig-rf-pred) showed strong alignment, although slight underprediction was observed in counties with extreme mortality rates, indicating potential underfitting at the distribution tails.

```{r}
#| label: tbl-model-performance
#| tbl-cap: "Test set performance metrics for Random Forest, LASSO regression, and a null model. RMSE = root mean squared error, MAE = mean absolute error, and R² = coefficient of determination. The null model predicts the mean mortality rate and serves as a baseline for comparison."
#| echo: false
#| #| fig-width: 9
#| fig-align: center
knitr::include_graphics(here::here("results", "tables", "final_model_performance_table.png"))
```

```{r}
#| label: fig-rf-pred
#| fig-cap: "Predicted versus observed mortality rates for the Random Forest model on the test set. Dashed line represents perfect prediction. Test set R² = 0.748."
#| echo: false
knitr::include_graphics(here::here("results", "figures", "rf_pred_vs_obs.png"))

@@

```{r}
#| label: fig-rf-vip
#| fig-cap: "Variable importance from the Random Forest model. Importance was calculated using permutation-based methods. Results are shown for the top 20 predictors."
#| echo: false
knitr::include_graphics(here::here("results", "figures", "rf_variable_importance.png"))

@@

The LASSO regression model served as a more interpretable, though slightly less accurate, alternative. On the test set, it achieved an RMSE of 119.5, MAE of 70.7, and R² of 0.726 (see @tbl-model-performance). The model applied coefficient shrinkage and automatic variable selection using a penalty parameter (λ = 1e–4), which was selected via cross-validation to minimize RMSE. This regularization excluded several dummy-coded predictors, improving interpretability by focusing on the most informative features. However, LASSO’s linear structure and shrinkage approach limited its flexibility in capturing non-linear interactions. The predicted versus observed plot (@fig-lasso-pred) displayed clear horizontal banding patterns, a result of many counties sharing identical categorical predictor combinations. These banding effects were more pronounced than in the Random Forest model, which better captured complex interactions through hierarchical tree splits. Supplementary results confirmed strong generalization in LASSO model performance across folds, with a mean RMSE of 119.63 (95% CI: 117.24–122.02) and R² of 0.73 (95% CI: 0.72–0.73) across 10-fold CV resamples (see Supplementary Table S4).

```{r}
#| label: fig-lasso-pred
#| fig-cap: "Predicted versus observed mortality rates for the LASSO regression model on the test set. Dashed line represents perfect prediction. Horizontal banding reflects repeated predictions due to dummy-encoded categorical variables."
#| echo: false
knitr::include_graphics(here::here("results", "figures", "lasso_pred_vs_obs.png"))

To contextualize model performance, we also evaluated a null model that predicted the mean training set mortality rate for all test observations. As expected, this baseline yielded substantially worse performance: RMSE = 228.65, MAE = 192.28, and an undefined R² due to zero variance in predictions (see Table 3). In contrast, both the Random Forest and LASSO models performed significantly better, confirming their ability to capture meaningful structure in the data. These results underscore that even with a limited set of demographic and temporal features, machine learning models can explain a substantial portion of the variation in county-level mortality.
